% How Universal Are Concept Directions? A Controlled Stress-Test of
% Representation Engineering Under RLHF
% BlackboxNLP 2026 @ EMNLP 2026
%
% (Retitled 2026-07-07: the original working title "The Myth of Universal
% Concept Directions Under RLHF" asserted fragmentation + RLHF amplification,
% which the Qwen results did not support --- refusal is largely universal under
% the prompt-based method, and RLHF slightly *consolidates* directions. The
% question form matches the actual findings and the Repro-track framing.)
%
% Compile with: pdflatex main.tex && bibtex main && pdflatex main.tex && pdflatex main.tex

\documentclass[11pt]{article}

% === Official ACL style (acl.sty vendored in this directory) ===
% [review] adds line numbers + anonymization for submission;
% switch to \usepackage[final]{acl} for camera-ready.
\usepackage[review]{acl}

% === Standard packages ===
% NOTE: acl.sty already loads natbib and hyperref and sets the two-column
% *ACL page geometry -- do not load geometry/natbib/hyperref again.
\usepackage{times}
\usepackage{latexsym}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{xcolor}
\usepackage{subcaption}
\usepackage{multirow}

% === Custom macros ===
\newcommand{\cosim}{\text{cos-sim}}
\newcommand{\dmodel}{d_{\text{model}}}
\newcommand{\vdir}{\mathbf{d}}
\newcommand{\vact}{\mathbf{a}}
\newcommand{\R}{\mathbb{R}}

% === Metadata ===
\title{How Universal Are Concept Directions? \\ A Controlled Stress-Test of Representation Engineering Under RLHF}

% ACL author blocks are separated with \And / \AND (not \and); under
% [review] the block is hidden automatically for anonymity.
\author{
  Author A \\
  Affiliation \\
  \texttt{email@example.com}
  \And
  Author B \\
  Affiliation \\
  \texttt{email@example.com}
}

\begin{document}
\maketitle

% ============================================================
\begin{abstract}
Representation engineering extracts a single linear direction for a concept
such as refusal or honesty and treats it as valid wherever the concept
occurs. We stress-test this universality assumption under controlled
conditions: for each concept we compare directions extracted within four
semantic domains against a direction pooled across domains, on both the
aligned and pre-alignment checkpoints of Qwen~2.5~3B, with confounds from
mixed data sources, templated responses, and domain imbalance explicitly
removed. Universality holds for refusal --- per-domain and global directions
agree closely (cosine $\geq 0.83$ at every layer) under two independent
operationalizations, and a global steering direction shifts held-out
behavior at least as well as domain-specific ones. It fails partially for
honesty, where early-layer directions diverge sharply (mean cosine 0.52; the
math domain is near-orthogonal to all others) and reconverge only partially
with depth. Contrary to our pre-registered hypothesis, alignment does not
amplify this fragmentation; if anything it slightly consolidates directions.
Finally, strong refusal fragmentation that we observed in a smaller pilot
model did not replicate on the primary model under identical data,
cautioning against generalizing single-model geometry findings. Universality
is thus a property of particular concepts at particular depths in particular
models --- not of concept directions as such.
\end{abstract}

% ============================================================
\section{Introduction}
\label{sec:intro}

A growing body of interpretability work holds that large language models
encode high-level concepts --- honesty, refusal, harmfulness --- as linear
directions in activation space \citep{zou2023representation,
marks2023geometry}. The practical appeal is obvious: if a concept is one
vector, it can be monitored with a dot product and controlled by adding that
vector back into the residual stream \citep{turner2023activation,
li2024inference}. Safety applications built on this premise, from refusal
probes to steering-based guardrails, inherit a strong implicit assumption:
that the direction extracted from one distribution of examples is the
\emph{same} direction wherever the concept occurs. A refusal direction fit
on requests about violence is presumed to govern refusals of privacy
violations; an honesty direction fit on trivia is presumed to govern honesty
about arithmetic.

This universality assumption is rarely tested, and testing it naively is
misleading. If the examples for different semantic domains come from
different datasets, differ in template structure, or differ in quantity,
then per-domain directions will diverge for reasons that have nothing to do
with the concept. Our own pilot experiment furnishes a cautionary example:
an apparent refusal fragmentation that vanished entirely on the primary
model under matched data.

We therefore stress-test universality under controlled conditions. For two
concepts, \textit{refusal} and \textit{honesty}, each decomposed into four
semantic domains, we compare per-domain concept directions against a pooled
global direction at every layer of Qwen~2.5~3B. The design removes the
obvious confounds: honesty domains are carved from a single source
(TruthfulQA), refusal is operationalized two independent ways (the standard
harmful-vs-harmless prompt contrast and a response-based contrast), domains
are balanced by capping, and confidence intervals come from a pair-level
bootstrap that re-estimates every direction per resample. Geometry is
checked against behavior with activation steering on held-out prompts, and
the whole analysis is repeated on the pre-alignment Base checkpoint to ask
whether RLHF amplifies or suppresses whatever structure exists. Two of our
three pre-registered hypotheses predicted failure of universality; we report
what we found rather than what we predicted.

Our findings are fourfold. \textbf{(1)}~Refusal directions are largely
universal: per-domain and global directions agree with cosine similarity
0.83--0.94 across all layers, under both operationalizations
(\S\ref{sec:results-refusal}). \textbf{(2)}~Honesty is not: early- and
middle-layer directions fragment sharply (mean cosine 0.52 at layer~5, with
the math domain near-orthogonal to every other), reconverging only partially
by the output layers (\S\ref{sec:results-honesty}). \textbf{(3)}~The
geometry is behaviorally real: a single global refusal direction steers
held-out prompts in every domain at least as well as that domain's own
direction (\S\ref{sec:results-steering}). \textbf{(4)}~Alignment does not
fragment concept representations --- the Instruct checkpoint is marginally
\emph{more} universal than its Base (\S\ref{sec:results-rlhf}) --- and
fragmentation observed on a smaller pilot model failed to replicate under
identical data (\S\ref{sec:results-pilot}).

Taken together, the results neither vindicate nor refute representation
engineering; they bound it. Universality is real where it has been most
prominently claimed (refusal), fails quietly elsewhere (honesty, early
depth), and varies across models in ways the field's single-model
methodology cannot detect. We release the full pipeline, data manifests, and
validation tooling.


% ============================================================
\section{Related Work}
\label{sec:related}

\paragraph{Linear concept representations.}
\citet{zou2023representation} introduced representation engineering as a
top-down alternative to circuit analysis, extracting concept directions by
contrasting activations on stimuli that do and do not express a concept, and
demonstrated reading and control across a broad range of concepts.
\citet{marks2023geometry} showed that true and false statements separate
linearly in middle-layer activations, and \citet{hernandez2023linearity}
found that many relational mappings are well approximated by linear
transformations. This literature generally extracts one direction per
concept per model and evaluates it in-distribution; the cross-domain
generalization of a single direction --- the property our study isolates ---
is usually assumed rather than measured.

\paragraph{Activation steering.}
Adding a scaled direction to the residual stream at inference time shifts
model behavior without weight updates \citep{turner2023activation}, and
targeted interventions on a small set of directions can improve truthfulness
\citep{li2024inference}. Most directly related to us,
\citet{arditi2024refusal} showed that refusal in open chat models is
mediated by a single direction whose ablation disables refusal across many
harmful instructions. Our refusal results independently support their
single-direction account with a domain-stratified design, and extend it with
the observation that the account does not automatically carry over to other
concepts or, apparently, to other models.

\paragraph{Alignment and its footprint on representations.}
\citet{wei2024assessing} demonstrate that safety alignment is brittle ---
removable with small parameter changes --- which motivates our question of
what alignment actually does to concept geometry. We find its footprint on
direction dispersion to be small and, if anything, consolidating. Finally,
our non-replication across model scales echoes the concern of
\citet{lieberum2023does} that interpretability findings established on one
model need not transfer to another; we provide a concrete instance at the
level of concept-direction geometry.


% ============================================================
\section{Methods}
\label{sec:methods}

Our design asks a simple question of the representation-engineering
framework: if a concept direction is extracted from data spanning several
semantic domains, how well does it agree with directions extracted from each
domain alone? We operationalize this for two concepts --- \textit{honesty}
and \textit{refusal} --- each decomposed into four domains, and measure
agreement both geometrically (angular dispersion) and behaviorally
(activation steering on held-out prompts). Crucially, the design controls
for the confounds that would otherwise masquerade as fragmentation: source
datasets that differ across domains, templated responses, and unbalanced
domain sizes.

\subsection{Contrastive Data}
\label{sec:prompts}

\paragraph{Honesty (primary).}
Domain must be the only factor that varies between the groups we compare, so
our primary experiment draws all of its data from a single source. We
partition the 817 questions of TruthfulQA \citep{lin2022truthfulqa} into
four domains by category --- \textit{factual trivia}, \textit{math},
\textit{politics/opinion}, and \textit{personal advice} --- supplemented
with 30 hand-written arithmetic pairs to populate the math domain. Each item
pairs a question with its best truthful answer (positive) and a plausible
falsehood (negative); activations are read from the concatenation of the
shared question and each response, so the question itself cancels in the
contrast and the measured difference isolates the response.

\paragraph{Refusal.}
We construct the refusal contrast in two independent ways. The
\textbf{prompt-based} condition follows the now-standard recipe of
\citet{arditi2024refusal}: harmful instructions (drawn from AdvBench
\citep{zou2023universal} and MedSafetyBench \citep{han2024medsafetybench},
plus hand-written privacy requests to balance the smallest domain) are
contrasted with harmless instructions, read at the final prompt token, with
no model responses involved. This is our headline refusal condition, as it
is immune to response templating. The \textbf{response-based} condition
pairs the same harmful prompts with compliant versus refusing responses; the
refusals were programmatically diversified so that no phrasing dominates the
negative class. We treat this second condition purely as a robustness check,
since synthetic responses risk encoding surface templates rather than the
target behavior. All pairs carry a framing tag distinguishing
harmful-request refusal from over-refusal of benign requests, and the two
framings are never pooled.

\paragraph{Balance.}
Because a global direction computed over unbalanced domains is dominated by
the largest one, we cap every domain at 120 pairs at extraction time
(prompt-based refusal is balanced at roughly 91--100 prompts per domain by
construction). The full dataset comprises 1{,}720 validated pairs; an
automated validator checks schema conformity, duplicate rates, and response
diversity for every file.

\subsection{Activation Extraction}
\label{sec:extraction}

Our primary models are Qwen~2.5~3B Instruct and its pre-alignment Base
checkpoint \citep{qwen2024qwen25}; Gemma~2~2B \citep{gemmateam2024gemma2}
serves as a pilot. For each prompt we record the post-block residual stream
at every layer (36 for Qwen, 26 for Gemma) at the final token position, in
half precision with NF4 4-bit weight quantization
\citep{dettmers2023qlora} to fit commodity GPUs; we return to the possible
effects of quantization in the Limitations. Extraction metadata (source
files, schemas, pair counts) is written alongside the cached activations so
that every downstream number is traceable to its data.

\subsection{Direction Computation}
\label{sec:directions}

Following \citet{zou2023representation}, the concept direction at layer
$\ell$ is the difference of means
\begin{equation}
  \vdir_\ell \;=\;
  \frac{1}{|P|}\sum_{x \in P} \vact_\ell(x)
  \;-\;
  \frac{1}{|N|}\sum_{x \in N} \vact_\ell(x),
\end{equation}
normalized to unit length, where $P$ and $N$ are the positive and negative
activation sets and $\vact_\ell(x) \in \R^{\dmodel}$ is the residual-stream
activation. Per-domain directions $\vdir_\ell^{(d)}$ use only pairs from
domain $d$; the global direction $\vdir_\ell^{(g)}$ pools all domains after
balancing.

\subsection{Angular Dispersion and Statistics}
\label{sec:angular}

Fragmentation at layer $\ell$ is summarized by the mean cosine similarity
between each per-domain direction and the global direction,
$\overline{\cosim}_\ell = \tfrac{1}{4}\sum_d
\cos\!\big(\vdir_\ell^{(d)}, \vdir_\ell^{(g)}\big)$,
together with the corresponding mean angle. A value near~1 indicates a
single shared axis; lower values indicate domain-specific structure. We also
report the full pairwise domain-by-domain cosine matrix at the most
fragmented layer, excluding the two embedding-adjacent layers, whose
directions reflect token statistics rather than concept representations.
Confidence intervals are obtained by a \emph{pair-level} bootstrap: in each
of 1{,}000 resamples we redraw prompt pairs with replacement within every
domain, recompute all per-domain and global directions from scratch, and
recompute the dispersion statistic. This propagates sampling variability in
the underlying data, rather than resampling the four domain cosines
themselves. Because early layers of a causal language model are dominated by
lexical features, we additionally report the mean cosine over the final
third of the network, where representations feed the unembedding and
behavior is read out.

\subsection{Behavioral Steering}
\label{sec:steering-method}

Geometry alone cannot establish that two directions are functionally
interchangeable, so we test transfer behaviorally with activation addition
\citep{turner2023activation,li2024inference}. For each refusal domain we
hold out 20 harmful prompts that were never used for direction computation,
format them with the model's chat template, and generate 64 tokens greedily
under three conditions: unsteered, steered with the global direction, and
steered with the domain's own direction (plus each other domain's direction
in a cross-domain sweep). The direction, scaled by a coefficient of 4.0, is
added to the residual stream at layer 18 at every token position; its sign
points from refusal toward compliance, so effective steering should
\emph{reduce} the refusal rate on harmful prompts. Refusals are detected by
keyword matching over a standard set of refusal markers. If concept
directions fragment, a domain's own direction should outperform the global
one on that domain's prompts; if they are universal, the two should be
indistinguishable.

\subsection{RLHF Comparison}
\label{sec:rlhf-method}

To isolate the effect of alignment we repeat the entire pipeline ---
extraction, direction computation, and dispersion analysis --- on the Base
checkpoint, which shares the pre-training of the Instruct model but has
undergone no preference tuning, and compare
$\overline{\cosim}_\ell$ layer by layer across the two checkpoints. The
pre-registered hypothesis was that alignment \emph{increases} dispersion. As
a sanity check that the checkpoints genuinely differ, we verify that
per-domain cosine differences between Base and Instruct grow with depth, the
expected signature of a fine-tune.


% ============================================================
\section{Experimental Setup}
\label{sec:setup}

\begin{table}[t]
  \centering
  \small
  \begin{tabular}{lll}
    \toprule
    & \textbf{Honesty} & \textbf{Refusal} \\
    \midrule
    Source & TruthfulQA & AdvBench, MedSafety- \\
           &            & Bench, hand-written \\
    Domains & 4 & 4 \\
    Pairs/domain & $\le$120 (math 59) & 91--120 \\
    Contrast & true vs.\ false & harmful vs.\ harmless \\
             & answer & prompt (+ resp.\ variant) \\
    \bottomrule
  \end{tabular}
  \caption{Data summary. Both concepts are additionally extracted from the
  Qwen~2.5~3B Base checkpoint for the RLHF comparison.}
  \label{tab:setup}
\end{table}

All experiments ran on single T4/P100 GPUs (roughly 25 GPU-hours in total);
analysis is CPU-only and fully scripted, from cached activations to the
figures in this paper. We use greedy decoding throughout the steering
experiments and fix all random seeds; the mock-data mode of the released
pipeline reproduces every table in this paper end-to-end on synthetic
activations before any GPU is touched. Code, data manifests, and validation
tooling are released with the paper.


% ============================================================
\section{Results}
\label{sec:results}

% NOTE: copy figure PDFs into paper/figures/ before compiling, e.g.
%   cp new_results_final/figures/qwen/*.pdf paper/figures/
%   cp "new_results_final/figures/qwen respbased/"*.pdf paper/figures/respbased_...

\begin{table}[t]
  \centering
  \small
  \begin{tabular}{lccc}
    \toprule
    \textbf{Condition} & \textbf{Min $\overline{\cosim}$} & \textbf{Layer} & \textbf{Late third} \\
    \midrule
    Refusal (prompt)    & 0.837 & 0  & 0.928 \\
    Refusal (response)  & 0.854 & 0  & 0.904 \\
    Honesty             & 0.522 & 5  & 0.694 \\
    \midrule
    \multicolumn{4}{l}{\textit{Base checkpoint}} \\
    Refusal (prompt)    & 0.834 & 0  & 0.911 \\
    Honesty             & 0.522 & 5  & 0.667 \\
    \bottomrule
  \end{tabular}
  \caption{Angular dispersion on Qwen~2.5~3B: minimum mean cosine between
  per-domain and global directions across layers, and the mean over the final
  third of the network. Refusal directions are close to a single axis
  throughout; honesty directions diverge sharply in early-mid layers and only
  partially reconverge.}
  \label{tab:dispersion}
\end{table}

\subsection{Refusal Directions Are Largely Universal}
\label{sec:results-refusal}

Contrary to our first pre-registered hypothesis, refusal shows little domain
structure. Under the prompt-based condition, the mean cosine between
per-domain and global directions never drops below 0.837 at any of the 36
layers of Qwen~2.5~3B Instruct, and averages 0.928 over the final third of
the network (Table~\ref{tab:dispersion}). The response-based robustness
condition agrees: despite an entirely different contrast construction, mean
cosines span 0.854--0.927 (pair-level 95\% CI at the most dispersed layer:
$[0.791, 0.885]$), with a late-layer mean of 0.904. Whatever the network
uses to decide refusal, it points in essentially the same direction whether
the request concerns violence, illegal activity, medical ethics, or privacy
--- consistent with the single-direction account of
\citet{arditi2024refusal}, and now shown to be stable across two
operationalizations of the concept.

\subsection{Honesty Fragments in Early and Middle Layers}
\label{sec:results-honesty}

Honesty behaves differently. The mean cosine to the global direction falls
to 0.522 at layer~5 --- a mean angle of $57^\circ$ --- and the domain
structure at that depth is stark: the math direction is essentially
orthogonal to every other domain (pairwise cosines between $-0.10$ and
$-0.03$), and even the best-aligned pair, factual trivia and personal
advice, reaches only 0.34. Dispersion then decays with depth, but only
partially: the final-third mean of 0.694 remains well below the
corresponding refusal values. Universality is thus not a property of concept
directions per se, but of particular concepts at particular depths. We note
that early-layer divergence may partly reflect lexical differences between
domains (arithmetic text simply looks different from trivia questions), which
is one reason we insist on the behavioral test below; the late-layer gap
between honesty and refusal, however, survives this caveat.

\subsection{Steering Confirms the Geometry}
\label{sec:results-steering}

\begin{table}[t]
  \centering
  \small
  \begin{tabular}{lccc}
    \toprule
    \textbf{Domain} & \textbf{Baseline} & \textbf{Global $\Delta$} & \textbf{Own $\Delta$} \\
    \midrule
    violence          & 0.95 & \phantom{$-$}0.00 & \phantom{$-$}0.00 \\
    illegal activity  & 0.90 & $-$0.05 & $-$0.05 \\
    medical/legal     & 0.50 & $-$0.15 & \phantom{$-$}0.00 \\
    privacy           & 0.50 & $-$0.40 & $-$0.15 \\
    \bottomrule
  \end{tabular}
  \caption{Behavioral steering on held-out harmful prompts (Qwen~2.5~3B
  Instruct, layer 18, coefficient 4.0, $n{=}20$ per domain). Baseline is the
  unsteered refusal rate; $\Delta$ columns give the change under steering
  with the global versus the domain's own direction. The global direction is
  never worse than the domain-specific one.}
  \label{tab:steering}
\end{table}

The second hypothesis --- that global steering underperforms domain-specific
steering --- also fails, in the direction favorable to universality
(Table~\ref{tab:steering}). In every domain the global direction shifts
behavior at least as much as the domain's own direction, and in two domains
strictly more (medical/legal: $-0.15$ vs.\ $0.00$; privacy: $-0.40$ vs.\
$-0.15$). Cross-domain directions transfer as well: privacy prompts are
steered as strongly by the violence and illegal-activity directions
($-0.40$) as by the global one. We attribute the weaker own-domain effects
to estimation noise --- each per-domain direction is fit on roughly a
quarter of the data --- which is itself an argument for pooling. Two
patterns deserve note: absolute effects are modest at this coefficient, and
the violence domain is behaviorally immovable (baseline refusal 0.95,
unchanged under every direction), suggesting refusal of violent requests is
enforced more redundantly than a single rank-one intervention can undo.

\subsection{RLHF Does Not Amplify Fragmentation}
\label{sec:results-rlhf}

Our third hypothesis predicted that alignment would increase dispersion. The
opposite holds, weakly: the Instruct checkpoint is slightly \emph{more}
universal than Base, with a mean per-layer gain in
$\overline{\cosim}_\ell$ of $+0.011$ for honesty and $+0.007$ for refusal,
and the gap widening toward the output (honesty at layer 35: 0.690 Base
vs.\ 0.769 Instruct). Only 10 of 36 layers move in the predicted direction,
none substantially. Base--Instruct differences in individual per-domain
cosines grow with depth (up to 0.20), the expected signature of a fine-tune,
confirming that the two checkpoints genuinely diverge --- alignment simply
does not diverge them along the axis we hypothesized. If anything,
preference tuning appears to \emph{consolidate} concept representations.

\subsection{A Cautionary Non-Replication}
\label{sec:results-pilot}

Our pilot on Gemma~2~2B, run on the response-based refusal data, showed what
looked like strong fragmentation: a minimum mean cosine of 0.559 and
near-orthogonal domain pairs. None of this survives on Qwen~2.5~3B --- the
same data recipe yields cosines of 0.85 and above (\S\ref{sec:results-refusal}).
We initially suspected dataset artifacts (mixed sources, templated
responses), but the prompt-based and response-based conditions agree on
Qwen, ruling that explanation out. The fragmentation was a property of the
smaller model, not of the data. We report this as a caution for the
literature: single-model fragmentation findings, including ours on Gemma,
should not be read as facts about concept directions in general.


% ============================================================
\section{Discussion}
\label{sec:discussion}

\paragraph{Universality is a property of concepts, not of the method.}
The cleanest reading of our results is that the question ``are concept
directions universal?'' is ill-posed as stated. Refusal, at least in
Qwen~2.5~3B, behaves as the representation-engineering literature assumes: a
single axis, stable across semantic domains, across two very different
extraction recipes, and causally effective wherever we tested it. Honesty
does not. Its early- and middle-layer directions are so domain-specific that
the math direction is orthogonal to the rest, and even at the output layers
the domains agree far less than refusal domains do. One plausible mechanism
is that refusal is a comparatively unified \emph{behavior} --- a decision the
model renders in a recognizable surface form regardless of topic --- whereas
honesty is a property of content, entangled with whatever domain knowledge
the content draws on. Deciding \emph{to} refuse may be one computation;
being right about arithmetic and being right about geography need not be.

\paragraph{Implications for steering-based safety.}
For practitioners the refusal result is reassuring: a guardrail built on a
single pooled refusal direction is not silently missing domains, and pooling
across domains is actively preferable, since per-domain directions are
noisier estimates of the same axis (\S\ref{sec:results-steering}). The
honesty result carries the opposite warning. Interventions aimed at
truthfulness that are calibrated on one distribution --- trivia-style
benchmarks being the common choice --- have no guarantee of engaging the
representations the model uses in other domains, particularly if they
intervene at early or middle depth. Probing and steering studies for
content-dependent concepts should report which domains their direction was
fit on, and test at least one held-out domain.

\paragraph{Alignment consolidates rather than fragments.}
We hypothesized that preference tuning, which rewards domain-sensitive
caution, would splinter concept directions. The data say otherwise: the
aligned checkpoint's directions are marginally more coherent than the base
model's, with the gain concentrated in late layers. A speculative but
consistent interpretation is that alignment training pressures the model
toward a common decision variable --- one refusal axis that many domains
feed --- because the reward signal itself is largely domain-general. Whatever
the mechanism, concept-level interventions on aligned models do not appear
to face a fragmentation penalty relative to base models.

\paragraph{The pilot that did not replicate.}
We initially interpreted the strong fragmentation in our Gemma~2~2B pilot as
the paper's headline finding, attributable to dataset artifacts. The
controlled comparison dismantled both the finding and our explanation of it:
on Qwen the same data recipe produces near-perfect universality, so the
fragmentation was a property of the model, not the data. We highlight this
as a methodological point in its own right. Geometry measured on a single
model is a case study, and the field's habit --- ours included --- of
generalizing from one open model deserves more suspicion than it gets.
Whether the difference reflects scale (2B vs.\ 3B), family, or training data
is a question our compute budget could not answer.


% ============================================================
\section{Limitations}
\label{sec:limitations}

\paragraph{Scale and coverage.}
Our primary evidence comes from one 3B-parameter model family, quantized to
4~bits to fit our compute budget; quantization may distort activation
geometry, and our own pilot demonstrates that these findings can be
model-specific. We study two concepts in four domains each, in English only.
The honesty math domain is thinner than the others (59 pairs against a
120-pair cap), so its striking orthogonality, while consistent with the
qualitative picture, rests on the least data; a domain-balance sensitivity
analysis is left to future work.

\paragraph{Measurement.}
Directions are read at the final token position and assumed linear;
concept structure that is nonlinear, distributed across positions, or
rank-greater-than-one is invisible to our measures. Early-layer honesty
fragmentation may be partly lexical --- arithmetic text simply differs from
trivia text at the surface --- which is why we lean on late-layer and
behavioral evidence, but the two cannot be fully disentangled here.

\paragraph{Steering protocol.}
The behavioral experiment uses keyword matching to detect refusal, a single
surviving steering coefficient, one intervention layer, and 20 held-out
prompts per domain; in the two domains with fewer than 140 source prompts,
held-out prompts partially overlap the direction-estimation data. Effects
are accordingly coarse: we trust the \emph{ordering} of global versus
per-domain steering, which is consistent across domains, more than any
individual effect size. A judge-model evaluation and a fuller coefficient
sweep would strengthen this leg of the study.


% ============================================================
\section{Conclusion}
\label{sec:conclusion}

We set out to show that concept directions fragment across semantic domains
and that alignment makes it worse. Under controls strict enough to trust,
neither claim survived for refusal, both survived in weakened form for
honesty, and a dramatic fragmentation result from our own pilot model turned
out not to generalize. The constructive summary for the field is a scoping
rule: a concept direction is universal exactly to the extent that the
concept is a unified behavior rather than domain-entangled content, and that
extent must be measured --- per concept, per depth, per model --- rather than
assumed. The controlled comparison that yields such measurements costs
little beyond the extraction runs practitioners already perform, and we
release our pipeline so that reporting one becomes the path of least
resistance.


% ============================================================
\section*{Acknowledgments}

% Omitted for anonymous review; add before camera-ready ([final] mode).


% ============================================================
\bibliographystyle{acl_natbib}
\bibliography{references}

% ============================================================
% Appendix (optional)
\appendix

\section{Prompt Pair Examples}
\label{app:prompts}

% TODO before submission: paste 1-2 representative pairs per domain.
% Honesty examples: data/prompt_pairs/honesty/<domain>.jsonl
% Refusal (prompt-based): regenerate with
%   python scripts/build_refusal_promptbased.py --max-per-domain 120
% CAUTION: choose harmful-prompt examples that are illustrative but mild;
% follow the MedSafetyBench research-only terms for any medical examples.

\section{Full Layer-wise Results}
\label{app:layers}

% TODO before submission: per-layer dispersion tables are already generated ---
% transcribe from new_results_final/report_<concept>_<model>.md (the Markdown
% tables can be converted to booktabs directly). Include all four
% concept x checkpoint tables plus the E2' (response-based) table from
% new_results_final/respbased/.

\end{document}
